### 2023-05-02 submission/manuscript.tex
Finished first full draft of Results

For the WNWN cluster analysis, sampling distributions for the two
treatment groups were generated using human fecal and ocean data
originally take from the GlobalPaterns dataset (L129) {[}XXXXX{]}. To
generate a fecal and ocean distribution, the authors included any
operational taxonomic unit (OTU) that appeared in more than one of the 4
fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by how
many of the 7 samples the OTUs were observed in followed by their total
abundance across all 7 samples (L139). From this sorted list they
identified the identifiers of the first 2000 OTUs (L66). Returning to
the 7 samples they selected the data for the corresponding 2000 OTUs and
pooled the OTU abundances of the fecal and ocean samples separately to
create two sampling distributions (L144, L159-160, L197-198). Next, the
fecal and ocean distributions were mixed in 8 different fractions to
generate two community types that differend by varying effect sizes
(i.e., 1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an
effect size of 1 generated a null model with no difference between the
treatment groups. To simulate the variation in sequencing depth across
the 80 samples, they normalized he number of sequences from each of the
26 samples in the GlobalPatterns dataset so that the median number of
sequences (N\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000,
5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
normalized sequencing depths to generate 80 sampling depths. From each
community type, they simulated 40 samples by sampling to the desired
number of sequences (L73, L230-233 and L326-327). Each simulation
condition was repeated 5 times (L85). This resulted in 160 simulations
(8 effect sizes x 4 median sampling depths x 5 replicates = 160
simulations). Finally, they removed rare and low prevalence OTUs in two
steps. First, they removed any OTUs whose total abundance was less than
3 across all 80 samples and that did not appear in at least 3 samples
(L368-386). Second, they removed any OTUs that did not have more than 1
sequence in more than 5\% of the 80 samples (i.e., 4 samples) and that
did not have a total abundance across the 80 samples greater than one
half of the number of samples in each community type (i.e., 20)
(L523-538, L551).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulated conditions were only replicated 5 times each
\item
  The average sizes of the libraries were small by modern standards
\item
  DESeq-based variance stabilization was used with distance calculation
  methods that are sensitive to negative values
\item
  A single subsampling of each dataset was evaluated rather than using
  rarefaction, which likely resulted in noisier data
\item
  Results using PAM clustering were not directly compared to those of
  K-means and hierarchical clustering
\item
  Subsampling removed the smallest 15\% of the samples, which penalized
  accuracy values by 15 percentage points
\item
  The distribution of library sizes was not typical of those commonly
  seen in microbiome analyses
\item
  A filtering step was applied to remove rare taxa from the simulated
  datasets
\item
  No accounting of case when library sizes are confounded with treatment
  group
\item
  Clustering accuracy was used rather than direct comparisons of beta
  diversity
\item
  There was no consideration of effects of normalization methods on
  alpha diversity metrics
\end{enumerate}

@@

Before assessing the impact of the points I critiqued above, I attempted
to replicate the results shown in Figures 4 and 5 of the original paper
using the authors original code and my own. I created a Conda
environment that used the R version and package versions that were as
close as possible to those used in the original paper. Because of the
slight differences in packages, it was necessary to apply several
patches to the original R-flavored markdown file to get the document to
render. I was able to generate a figure similar to that presented as
Figures 4 and 5 of the original paper. My results are shown in Figures
S2 (\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_3.pdf})
and S3
(\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_4.pdf}) of
this paper, respectively. The differences in results are likely due to
differences in software versions and operating systems. It is also worth
noting that their versions of the two figures differ from those included
in Protocol S1 within the rendered html file
(\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
and that the figure numbers are one higher in the paper than those
generated by the R-flavored markdown file. Regardless of the
differences, the results are qualitatively similar.

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reports the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in the original Figures 4 and 5 (e.g.~the
Bray-Curtis distances calculated on the DESeqVS normalized data). A
better approach would have been to use 100 replicates as this would
reduce the dependency of the results on the random number generator's
seed. By increasing the number of replicates it was also possible to
compare the probability of falsely and correctly clustering samples from
the same and different treatment groups together (see points 9-11,
below). Because the accuracies are unlikely to be symmetric around a
mean at larger accuracy values the median and 95\% confidence intervals
or intraquartile range should have been reported. To test the effect of
increasing the number of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment as above. This was necessary
since the number of simulated conditions because of this change
increased 20-fold. Such intense data processing was not practical within
a single R-flavored markdown document. Again, the observed results were
qualitiatively similar to those generated using the single R-flavored
markdown file (Figures S4 (\textbf{pam\_subsample15\_fig\_4.pdf}) and
S5(\textbf{pam\_subsample\_fig\_5.pdf}) ). The increased number of
replications resulted in smoother lines and allowed me to present
empirical 95\% confidence intervals. For all analyses in the remainder
of this paper, I used 100 randomized replicates per condition.

@@

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. For more
modern datasets, it would be reasonable to expect a median number of
sequences larger than 10,000 (see Table 1 of {[}Singleton Paper
XXXXXXXX{]}). Therefore, I included an additional average depth of
sampling value of 50,000 sequences with the four average sequencing
sampling depths as WNWN (i.e., 1,000, 2,000, 5,000, 10,000). Additional
sequencing coverage would be expected to result in more robust distance
values since there would be more information represented in the data.
Indeed, the added sampling depth showed higher accuracy values at lower
effect sizes for the combinations of normalization methods and distance
calculations (Figure S4 (\textbf{pam\_subsample15\_fig\_4.pdf})).
Increased sequencing coverage also resulted in improved clustering
accuracy for lower effect sizes when the library size minimum quantile
was decreased (Figure S5 (\textbf{pam\_subsample\_fig\_5.pdf})). I will
revisit the choice of the library size minimum quantile below.

@@

Close comparison of the original Figure 4 and my version
(\textbf{pam\_subsample15\_fig\_4.pdf}) revealed one important
difference between the two plots. In the original analysis, the
accuracies for the Weighted UniFrac distances at the largest effect size
(i.e., 3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and
10,000. In my version of the analysis, the values for the same
sequencing depths were 0.88, 0.89, and 1.00, respectively. The 95\%
confidence interval for these accuracies spaned 0.51 and 1.00. The
Bray-Curtis distances were also different by both methods at smaller
effect sizes and had wide confidence intervals. Inspection of the DESeq
normalized OTU counts revealed that the method resulted in negative
values. In fact, rendering the R-flavored markdown files in WNWN's
Protocol S1 generated a warning message when passing the DESeq
normalized counts to the Bray-Curtis calculator, which said, ``results
may be meaningless because data have negative entries in method
`bray'\,''. Although the weighted UniFrac calculator function did not
generate a similar warning message, negative values would also result in
similarly meaningless distances. Both are due to the fact that the
calculators sum the counts of each OTU in both samples being compared.
In contrast, a Euclidean distance does not use a similar sum, but sums
the square of the difference between the OTU abundance in each sample.

To assess the prevalence of negative counts in the simulated data, I
quantified the fraction of negative values in the OTU matrix from each
simulation and counted the number of simulations where the normalized
OTU table had at least one negative value (Figure
\textbf{deseq\_negative\_value.pdf}). In general the fraction of
negative OTU counts increased with effect size, but decreased with
sequencing effort. The fraction of simulations with at least onee
negative value increased with effect size and sequencing effort. The
high frequency of negative OTU counts resulted in highly variable
Bray-Curtis and weighted UniFrac values. It is likely that because the
WNWN analysis only used 5 replicates that the large variation in
accuracies at high effect sizes was missed initially. For the rest of
this reanalysis study, I will only report results using the DESeq-based
variance stabilization normalization with the Euclidean distance.

@@

As noted above, the original jargon that was used in WNWN was confusing
to many who conflated rarefying with rarefaction. One problem with a
single subsampling of a community is that it is unlikely to obtain a
representative sampling of each community. A more robust analysis would
have used rarefaction rather than a single subsampling of the data since
it would have averaged across random subsamplings, which individually
would be unlikely to represent the overall composition of the
communities. Rather than being guilty of ``omission of available valid
data'', rarefaction uses all of the available data. To compare
subsampling and rarefaction, I removed the 15\% of samples with the
lowest number of sequences for each of the 100 simulated datasets and
compared the distances from a single subsampling to rarefaction using
the distance calculations shown in the original Figure 4. This analysis
revealed two benefits of rarefaction. First, the median distances
generated by rarefaction was always as large or larger than those from a
single subsample (Figure \textbf{subsample\_rarefaction\_compare.pdf}).
In general, the difference was most pronounced for smaller average
library sizes and at smaller effect sizes. The unweighted UniFrac
distances were most impacted by the use of rarefaction over subsampling.
Second, the intraquartile ranges for the distances generated by
rarefaction were generally smaller than those by subsampling and showed
similar trends to the difference in the median distances (Figure
\textbf{subsample\_rarefaction\_compare.pdf}). The intraquartile ranges
for Bray-Curtis, Euclidean, and unweighted UniFrac distances were
actually larger by rarefaction than by subsampling at small effect sizes
and average library sizes; however at larger values the intraquartile
range by subsampling was larger than by rarefaction for these distance
calculations. Because rarefaction incorporates more of the data and
generally performed better than subsampling, the remainder of this
analysis will report results using rarefaction rather than by
subsampling except when noted.

@@

The clustering accuracy measurements in the body of the manuscript were
determined using PAM-based clusters while Protocol S1 also includes
K-means and hierarchichal clustering. Although the data were not
displayed in a manner that lent itself to direct comparison, close
inspection of the rendered figures suggests that PAM may not have been
the optimal choice in all situations. Rather, K-means clustering may
have been prefered. Because the accuracies were the smallest at lower
effect sizes, I focused my comparison at the effect size of 1.15. For
each set of 100 replicated simulated datasets, I compared the clustering
accuracy across clustering methods to see how often each clustering
method resulted in the highest accuracy (Figure
\textbf{compare\_cluster\_methods.pdf}). Indeed, K-means clustering
performed better than the other methods. Among all combinations of
normalization methods, distance calculations, and read depths, PAM
clustering resulted in clustering accuracies as good or better than the
other methods in 49.92\% of the randomizations (Figure
\textbf{compare\_cluster\_methods.pdf}). K-means clustering was at least
as good as the other methods in 74.39\% of the randomizations. HClust
was at least as good as the other methods for 44.32\% of the
randomizations. I specifically compared the clustering accuracies using
rarefaction for each of the distance calculations methods using PAM and
K-means clustering. Among the 30 combinations of distance calculations
and read depths, K-means performed better than PAM in 29 cases with PAM
doing better in the 1 other case when calculating distances with
Euclidean using 10,000 sequences. When using subsampled data, K-means
clustering performed better than PAM in each case. Because K-means
clustering did so much better than PAM clustering in the simulated
conditions, I will use K-means clustering for the remainder of this
study.

@@

In WNWN, the authors quantified the tradeoff between sampling effort,
the number of samples removed below the threshold, and clustering
accuracy (original Figure 5, my Figure S5). Although the optimal
sampling effort varied by distance metric, normalization method, and
sampling effort, they removed samples whose number of sequences was less
than the 15th percentile (L404-419). They acknowledged that this
screening step, which was only used with subsampling, would decrease
clustering accuracy putting it at a relative disadvantage to the other
methods (page 5, column 1, last paragraph). Therefore, it was not
surprising that the peak clustering accuracy for their subsampled data
was at 85\% (Figure S2). Because the true result would not known \emph{a
priori} in microbiome studies, it would be impossible for researchers to
conduct a sensitivity analysis comparing the tradeoffs between
sequencing depth, sample number, and clustering accuracy to select a
sampling depth. {[}\textbf{Although the authors claim that ``Rarefying
counts requires an arbitrary selection of a library size minimum that
affects downstream inference'' (page 8, column 1, point 3), in acutal
microbiome studies the selection of a sampling depth is not as arbitrary
as the authors claim. Rather, to avoid p-hacking, researchers pick a set
of criteria where they will include or exclude samples prior to testing
their data.}{]} The differences in clustering accuracy between
subsampling and rarefaction and using PAM and K-means clustering
indicated that it was necessary to reassess the tradeoff between the
library size minimum quantile and clustering accuracy. When using
rarefaction, K-means clustering, and only considering conditions with
2,000 or more sequences, there was not a condition where setting a
higher threshold resulted in a better accuracy than using all of the
samples (\textbf{kmeans\_rarefaction\_fig\_5.pdf}). These results showed
that for modern sequencing depths, using the full datasets with
rarefaction and K-means clustering resulted in accuracies that were
typically better than those observed when removing the smallest 15\% of
the samples from each simulated dataset. When the original Figure 4 was
recast with these approaches, Rarefaction performed at least as well as
any of the other tranformations with each distance calculation, except
for with the Poisson distance (Figure
\textbf{kmeans\_rarefaction15\_fig\_4.pdf}). It is worth noting that at
larger effect sizes, K-means clustering did not perform as well for some
combinations of normalization methods and distance calculations (compare
\textbf{kmeans\_rarefaction15\_fig\_4.pdf} and
\textbf{pam\_subsample00\_fig\_4.pdf}); however, those combinations that
performed worse by K-means were not as good as rarefaction or
subsampling by either clustering method.

@@

As described above, the sequencing depths used in the 26 GlobalPatterns
datasets were used as the distribution to create sequencing depths for
the 80 samples that were generated in each simulation. The
GlobalPatterns datasets had a mean of 1,085,256.85 sequences and a
median of 1,106,849.00 sequences per dataset. The datasets ranged in
sequencing depth between 58,688.00 and 2,357,181.00 sequences for a
40.16-fold difference. Rather than representing a typically observed
distribution of sequencing depths that would be skewed right, the
sampling distribution was normally distributed (Shapiro-Wilk test of
normality, P=0.57) (Figure \textbf{\emph{distribution\_shape.pdf}}).
From these simulations it is unclear how sensitive the various
normalizations and distance calculations are to a skewed distribution. A
second limitation of this sampling distribution is that it only
contained 26 unique sampling depths such that each sampling depth would
have been re-used an average of 3.08 times in each simulation. Yet, it
is unlikely for a real sequence collection to have duplicate sequencing
depths.

I created a new set of simulations that would focus on solely on the
effect of the shape of the distribution on the results to reassess the
WNWN results in the context of a more typical distribution of sample
sizes. I created a simple sequencing depth distribution where there were
80 depths logarithmically distributed between the minimum and maximum
sequencing depths of the GlobalPatterns dataset (Figure
\textbf{\emph{distribution\_shape.pdf}}). The median of this
distribution was 372,040.00 and the mean was 629,825.00. When I
regenerated the original Figures 4 and 5 using the log-distributed
sequencing effort distribution, the differences in normalization methods
were more apparent (Figure
\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_log.pdf}}). For each of the
distance calculators, rarefying the data to the size of the smallest
dataset yielded accuracies that were at least as good as the other
methods across effect sizes and median sequencing depths. The difference
was most pronounced at smaller effect sizes and sequencing depths. When
comparing the performance of rarefaction across distance calculators for
different effect sizes, sequencing depths and size of smallest sample
(\textbf{\emph{fig\_5\_kmeans\_rarefaction\_log.pdf}}), the accuracies I
observed using the log-distributed sample sizes was at least as good as
those obtained using the GlobalPatterns-based distribution
\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_a.pdf}}. The issue of
number of samples and the distribution of their sequencing depths in the
context of controlling for uneven sampling effort is explored in far
greater detail in an another analysis using sequencing depths observed
in biological samples {[}XXXXXX{]}.

@@

McMurdie and Holmes were emphatic that ``\textbf{rarefying biological
count data is statistically inadmissible} because it requires the
omission of available valid data'' (emphasis in original). Thus it is
strange that they argue against removing data when
rarefying/subsampling, but accept removing rare and low-prevalence OTUs
prior to normalizing and analyzing their simulated communities. This
practice has become common in microbiome studies and is the standard
approach in tools such as dada2, unoise, and deblur {[}XXXXXXXX{]}.
However, my previous work has shown that rare sequences from a poorly
sequenced sample often appear in more deeply sequened samples suggesting
that they are not necessarily artifacts. Furthermore, removing rare
sequences alters the structure of communities and has undesirable
effects on downstream analyses {[}XXXXXXXX{]}.

>>>>

For the WNWN cluster analysis, sampling distributions for the two
treatment groups were generated using human fecal and ocean data
originally take from the GlobalPaterns dataset (L129) {[}XXXXX{]}. To
generate a fecal and ocean distribution, the authors included any
operational taxonomic unit (OTU) that appeared in more than one of the 4
fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by how
many of the 7 samples the OTUs were observed in followed by their total
abundance across all 7 samples (L139). From this sorted list they
identified the identifiers of the first 2000 OTUs (L66). Returning to
the 7 samples they selected the data for the corresponding 2000 OTUs and
pooled the OTU abundances of the fecal and ocean samples separately to
create two sampling distributions (L144, L159-160, L197-198). Next, the
fecal and ocean distributions were mixed in 8 different fractions to
generate two community types that differend by varying effect sizes
(i.e., 1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an
effect size of 1 generated a null model with no difference between the
treatment groups. To simulate the variation in sequencing depth across
the 80 samples, they normalized he number of sequences from each of the
26 samples in the GlobalPatterns dataset so that the median number of
sequences (Ñ\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000,
5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
normalized sequencing depths to generate 80 sampling depths. From each
community type, they simulated 40 samples by sampling to the desired
number of sequences (L73, L230-233 and L326-327). Each simulation
condition was repeated 5 times (L85). This resulted in 160 simulations
(8 effect sizes x 4 median sampling depths x 5 replicates = 160
simulations). Finally, they removed rare and low prevalence OTUs in two
steps. First, they removed any OTUs whose total abundance was less than
3 across all 80 samples and that did not appear in at least 3 samples
(L368-386). Second, they removed any OTUs that did not have more than 1
sequence in more than 5\% of the 80 samples (i.e., 4 samples) and that
did not have a total abundance across the 80 samples greater than one
half of the number of samples in each community type (i.e., 20)
(L523-538, L551).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulated conditions were only replicated 5 times each
\item
  The average sizes of the libraries were small by modern standards
\item
  DESeq-based variance stabilization was used with distance calculation
  methods that are sensitive to negative values
\item
  A single subsampling of each dataset was evaluated rather than using
  rarefaction, which likely resulted in noisier data
\item
  Results using PAM clustering were not directly compared to those of
  K-means and hierarchical clustering
\item
  Subsampling removed the smallest 15\% of the samples, which penalized
  accuracy values by 15 percentage points
\item
  The distribution of library sizes was not typical of those commonly
  seen in microbiome analyses
\item
  A filtering step was applied to remove rare taxa from the simulated
  datasets
\item
  No accounting for difference in performance when library sizes are
  confounded with treatment group
\item
  Clustering accuracy was used rather than direct comparisons of beta
  diversity
\item
  There was no consideration of effects of normalization methods on
  alpha diversity metrics
\end{enumerate}

@@

Before assessing the impact of the points I critiqued above, I attempted
to replicate the results shown in Figures 4 and 5 of the original paper
using the authors' original code. I created a Conda environment that
used the R version and package versions that were as close as possible
to those used in the original paper. Because of the slight differences
in packages, it was necessary to apply several patches to the original
R-flavored markdown file to render document. I was able to generate a
figure similar to that presented as Figures 4 and 5 of the original
paper. My results are shown in Figures S2
(\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_3.pdf}) and
S3 (\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_4.pdf})
of this paper, respectively. The differences in results are likely due
to differences in software versions and operating systems. It is also
worth noting that the published versions of the two figures differ from
those included in Protocol S1 within the rendered html file
(\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
and that the figure numbers are one higher in the paper than those
generated by the R-flavored markdown file (i.e., Protocol S1 has Figures
3 and 4 corresponding to the published Figures 4 and 5). Regardless of
the differences, my results were qualitatively similar to the originals.

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reports the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in the original Figures 4 and 5 (e.g.~the
Bray-Curtis distances calculated on the DESeqVS normalized data). A
better approach would have been to use 100 replicates as this would
reduce the dependency of the results on the random number generator's
seed. By increasing the number of replicates it was also possible to
compare the probability of falsely and correctly clustering samples from
the same and different treatment groups together (see points 9-11,
below). Because the accuracies were not symmetric around the mean
accuracy values the median and 95\% confidence intervals or
intraquartile range should have been reported. To test the effect of
increasing the number of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment I used above. This was
necessary since the number of simulated conditions increased 20-fold
with the additional replicates. Such intense data processing was not
practical within a single R-flavored markdown document. Again, the
observed results were qualitiatively similar to those generated using
the single R-flavored markdown file (Figures
\textbf{pam\_subsample15\_fig\_4.pdf} and
\textbf{pam\_subsample\_fig\_5.pdf}). The increased number of
replications resulted in smoother lines and allowed me to present
empirical 95\% confidence intervals. For all analyses in the remainder
of this paper, I used 100 randomized replicates per condition.

@@

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. For more
modern datasets, it would be reasonable to expect a median number of
sequences larger than 10,000 (see Table 1 of {[}Singleton Paper
XXXXXXXX{]}). Therefore, I included an additional median depth of
sampling value of 50,000 sequences with the original four median
sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000).
Additional sequencing coverage would be expected to result in more
robust distance values since there would be more information represented
in the data. Indeed, the added sampling depth showed higher accuracy
values at lower effect sizes for the combinations of normalization
methods and distance calculations (Figure S4
(\textbf{pam\_subsample15\_fig\_4.pdf})). Increased sequencing coverage
also resulted in improved clustering accuracy for lower effect sizes
when the library size minimum quantile was decreased (Figure S5
(\textbf{pam\_subsample\_fig\_5.pdf})). I will revisit the choice of the
library size minimum quantile below.

@@

Close comparison of the original Figure 4 and my version
(\textbf{pam\_subsample15\_fig\_4.pdf}) revealed one important
difference between the two plots. In the original analysis, the
accuracies for the Weighted UniFrac distances at the largest effect size
(i.e., 3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and
10,000. In my version of the analysis, the values for the same
sequencing depths were 0.88, 0.89, and 1.00, respectively. The 95\%
confidence interval for these accuracies spaned between 0.51 and 1.00.
The Bray-Curtis distances were also different by both methods at smaller
effect sizes and had wide confidence intervals. Inspection of the DESeq
normalized OTU counts revealed that the method resulted in negative
values. In fact, rendering the R-flavored markdown files in WNWN's
Protocol S1 generated warning messages when passing the DESeq normalized
counts to the Bray-Curtis calculator, which said, ``results may be
meaningless because data have negative entries in method `bray'\,''.
Although the Weighted UniFrac calculator function did not generate a
similar warning message, negative values would also result in similarly
meaningless distances. Both are due to the fact that the calculators sum
the counts of each OTU in both samples being compared. In contrast, a
Euclidean distance does not use a similar sum, but sums the square of
the difference between the OTU abundance in each sample. To assess the
prevalence of negative counts in the simulated data, I quantified the
fraction of negative values in the OTU matrix from each simulation and
counted the number of simulations where the normalized OTU table had at
least one negative value (Figure \textbf{deseq\_negative\_value.pdf}).
In general the fraction of negative OTU counts increased with effect
size, but decreased with sequencing effort. The fraction of simulations
with at least onee negative value increased with effect size and
sequencing effort. The high frequency of negative OTU counts resulted in
highly variable Bray-Curtis and Weighted UniFrac values. It is likely
that because the WNWN analysis only used 5 replicates that the large
variation in accuracies at high effect sizes was missed initially. For
the rest of this reanalysis study, I will only report results using the
DESeq-based variance stabilization normalization with the Euclidean
distance.

@@

As noted above, the original jargon that was used in WNWN was confusing
to many who conflated rarefying/subsampling with rarefaction. A more
robust analysis would have used rarefaction since it would have averaged
across random subsamplings, which individually would be unlikely to
represent the overall composition of the communities. Rather than being
guilty of ``omission of available valid data'' as claimed in WNWN,
rarefaction uses all of the available data. To compare subsampling and
rarefaction, I removed the 15\% of samples with the lowest number of
sequences for each of the 100 simulated datasets and compared the
clustering accuracies from a single subsampling to rarefaction with 100
randomizations. This analysis revealed two benefits of rarefaction.
First, the median distances generated by rarefaction was always at least
as large as those from a single subsample (Figure
\textbf{subsample\_rarefaction\_compare.pdf}). The difference was most
pronounced for smaller average library sizes and at smaller effect
sizes. The Unweighted UniFrac distances were most impacted by the use of
rarefaction over subsampling. Second, the intraquartile ranges for the
distances generated by rarefaction were generally smaller than those by
subsampling and showed similar trends to the difference in the median
distances (Figure \textbf{subsample\_rarefaction\_compare.pdf}). The
intraquartile ranges for Bray-Curtis, Euclidean, and Unweighted UniFrac
distances were actually larger by rarefaction than by subsampling at
small effect sizes and average library sizes; however at larger values
the subsampling intraquartile range was larger than by rarefaction for
these distance calculations. Because rarefaction incorporates more of
the data and generally performed better than subsampling, the remainder
of this analysis will report results using rarefaction rather than by
subsampling, except when noted.

@@

The clustering accuracy measurements in the body of the manuscript were
determined using PAM-based clusters while Protocol S1 also includes
K-means and hierarchichal clustering. Although the data were not
displayed in a manner that lent itself to direct comparison in Protocol
S1, close inspection of the rendered figures suggested that PAM may not
have been the optimal choice in all situations. Rather, K-means
clustering appeared to perform better in many simulations. Because the
accuracies were the smallest at lower effect sizes, I focused my
comparison at the effect size of 1.15. For each set of 100 replicated
simulated datasets, I compared the clustering accuracy across clustering
methods to see how often each clustering method resulted in the highest
accuracy (Figure \textbf{compare\_cluster\_methods.pdf}). Indeed,
K-means clustering performed better than the other methods. Among all
combinations of normalization methods, distance calculations, and read
depths, PAM clustering resulted in clustering accuracies as good or
better than the other methods in 49.92\% of the randomizations (Figure
\textbf{compare\_cluster\_methods.pdf}). K-means clustering was at least
as good as the other methods in 74.39\% of the randomizations. HClust
was at least as good as the other methods for 44.32\% of the
randomizations. I specifically compared the clustering accuracies using
rarefaction for each of the distance calculations methods using PAM and
K-means clustering. Among the 30 combinations of distance calculations
and read depths, K-means performed better than PAM in 29 cases with PAM
doing better in the 1 other case (i.e., calculating distances with
Euclidean using 10,000 sequences). When using subsampled data, K-means
clustering performed better than PAM in each case. Because K-means
clustering did so much better than PAM clustering in the simulated
conditions, I will use K-means clustering for the remainder of this
study.

@@

In WNWN, the authors quantified the tradeoff between median sequencing
depth, the number of samples removed below the threshold, and clustering
accuracy (original Figure 5, my Figure S5). Although the optimal
threshold varied by distance metric, normalization method, and
sequencing depth, they removed samples whose number of sequences was
less than the 15th percentile (L404-419). They acknowledged that this
screening step, which was only used with subsampling, would decrease
clustering accuracy putting it at a relative disadvantage to the other
methods (page 5, column 1, last paragraph). Therefore, it was not
surprising that the peak clustering accuracy for their subsampled data
was at 85\%. Because the true best result would not known \emph{a
priori} in an actual microbiome study, it would be impossible for
researchers to conduct a sensitivity analysis comparing the tradeoffs
between sequencing depth, sample number, and clustering accuracy to
select a sampling depth for their analysis. The differences in
clustering accuracy between subsampling and rarefaction and using PAM
and K-means clustering indicated that it was necessary to reassess the
tradeoff between the library size minimum quantile and clustering
accuracy. When using rarefaction, K-means clustering, and only
considering conditions with 2,000 or more sequences, there was not a
condition where setting a higher threshold resulted in a better accuracy
than using all of the samples
(\textbf{kmeans\_rarefaction\_fig\_5.pdf}). These results showed that
for modern sequencing depths, using the full datasets with rarefaction
and K-means clustering resulted in accuracies that were typically better
than those observed when removing the smallest 15\% of the samples from
each simulated dataset. When the original Figure 4 was recast with these
approaches, rarefaction performed at least as well as any of the other
tranformations with each distance calculation, except when used with the
Poisson distance (Figure \textbf{kmeans\_rarefaction15\_fig\_4.pdf}). It
is worth noting that at the largest effect sizes, K-means clustering did
not perform as well as PAM for some combinations of normalization method
and distance calculation (compare
\textbf{kmeans\_rarefaction15\_fig\_4.pdf} and
\textbf{pam\_subsample00\_fig\_4.pdf}); however, those combinations that
performed worse by K-means were not as good as rarefaction or
subsampling by either clustering method.

@@

As described above, the sequencing depths used in the 26 GlobalPatterns
datasets were used as the distribution to create sequencing depths for
the 80 samples that were generated in each simulation. The
GlobalPatterns datasets had a mean of 1085256.8 sequences and a median
of 1,106,849 sequences per dataset. The datasets ranged in sequencing
depth between 58,688 and 2,357,181 sequences for a 40.16-fold
difference. Rather than representing a typically observed distribution
of sequencing depths that would be skewed right, the sampling
distribution was normally distributed (Shapiro-Wilk test of normality,
P=0.57) (Figure \textbf{\emph{distribution\_shape.pdf}}). From these
simulations it is unclear how sensitive the various normalizations and
distance calculations were to a skewed distribution. A second limitation
of this sampling distribution is that it only contained 26 unique
sampling depths such that each sampling depth would have been re-used an
average of 3.08 times in each simulation. Yet, it is unlikely for a real
sequence collection to have duplicate sequencing depths. To reassess the
WNWN results in the context of a more typical distribution of sample
sizes, I created a new set of simulations to test the effect of the
shape of the distribution on the results. I created a simple sequencing
depth distribution where there were 80 depths logarithmically
distributed between the minimum and maximum sequencing depths of the
GlobalPatterns dataset (Figure \textbf{\emph{distribution\_shape.pdf}}).
The median of this distribution was 372,040 and the mean was 629824.8.
When I regenerated the original Figures 4 and 5 using the
log-distributed sequencing effort distribution, the differences in
normalization methods were more apparent (Figure
\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_log.pdf}}). For each of the
distance calculators, rarefaction to the size of the smallest dataset
yielded accuracies that were at least as good as the other methods
across effect sizes and median sequencing depths. The difference was
most pronounced at smaller effect sizes and sequencing depths. When
comparing the performance of rarefaction across distance calculators for
different effect sizes, sequencing depths and size of smallest sample
(\textbf{\emph{fig\_5\_kmeans\_rarefaction\_log.pdf}}), the accuracies I
observed using the log-distributed sample sizes was at least as good as
those obtained using the GlobalPatterns-based distribution
\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_a.pdf}}. The issue of the
number of samples in a study and the distribution of their sequencing
depths in the context of controlling for uneven sampling effort is
explored in far greater detail in an another analysis using sequencing
depths observed in actual biological samples {[}XXXXXX{]}.

@@

McMurdie and Holmes were emphatic that ``\textbf{rarefying biological
count data is statistically inadmissible} because it requires the
omission of available valid data'' (emphasis in original). Thus it is
strange that they argue against removing data when
rarefying/subsampling, but accept removing rare and low-prevalence OTUs
prior to normalizing their counts. This practice has become common in
microbiome studies and is the standard approach in tools such as dada2,
unoise, and deblur {[}XXXXXXXX{]}. However, my previous work has shown
that rare sequences from a poorly sequenced sample often appear in more
deeply sequened samples suggesting that they are not necessarily
artifacts. Furthermore, removing rare sequences alters the structure of
communities and has undesirable effects on downstream analyses
{[}XXXXXXXX{]}. Although my previous work does an extensive analysis of
the effects of removing rare sequences, I wanted to explore the effect
of filtering in the context of the WNWN simulation framework. For each
of the filtered and non-filtered OTU tables I calcualted the absolute
value of the difference in accuracy between each distance calculation
following the normalization procedure (Figure
\textbf{compare\_filter\_accuracy.pdf}). With the exception of the
Weighted UniFrac distances, each of the distance calculations and
normalization procedures were sensitive to the filtering. The
rarefaction data tended to be sensitive to filtering at small effect
sizes. Distances generated using raw counts, DESeq variance
stabilization, and Upper Quartile Log Fold Change tended to be more
sensitive to filtering at larger effect sizes. When using relative
abundance data, Bray-Curtis distances were sensitive to filtering at
small effect sizes and Unweighted UniFrac distances were sensitive at
large effect sizes. These trends appeared to be driven by the dependence
of the distance calculation on low abundance taxa. More surprising than
the effect of filtering on the mean absolute difference in clustering
accuracy was the wide variation in accuracies at each effect size. Among
the different normalization methods, the accuracies calculated using
rarefaction had the narrowest 95\% confidence interval for all distance
calculations except for calculating Unweighted UniFrac distances. For
these distances, using relative abundances had the narrowest range at
small effect sizes; rarefaction had the narrowest range at larger effect
sizes. Again, these trends appeared to be driven by the dependence on
low frequency taxa in Unweighted UniFrac, which is dependent on the
presence or absence of taxa rather than their abundance. Given my
previous work and the large variation caused by removing rare taxa, OTU
filtering should not be performed in microbiome analyses.

\hypertarget{no-accounting-for-difference-in-performance-when-library-sizes-are-confounded-with-treatment-group}{%
\subsubsection{9. No accounting for difference in performance when
library sizes are confounded with treatment
group}\label{no-accounting-for-difference-in-performance-when-library-sizes-are-confounded-with-treatment-group}}
